### 2017-01-23 doc/manuscript.Rmd
Second draft of discussion

- General findings
- Limitations of study
- Previous lit and limits of those
- How this study fills those gaps and new things learned
- Potential clinical application

@@

We found that in an at-risk for diabetes Canadian population of mainly female
and European ancestry, the specific TAGFA composition was strongly associated
with insulin sensitivity and moderately associated with beta-cell function. In
particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm }}
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. {{confirm}} While most TAGFA
were not associated with beta-cell function, {{confirm after checking models}}
two fatty acids, 16:0 and 18:1n-7, were associated negatively and positively,
respectively, with both measures of beta-cell function. We also found that four
TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7) clustered together, and that also strongly 
predicted lower insulin sensitivity. These four fatty acids are also involved in
the DNL of refined and simple carbohydrates {{cite}}. Our results, which are in
agreement with the current literature, suggest that higher intakes of simple
carbohydrates or higher activity of DNL may increase the risk for diabetes,
likely through worsening insulin sensitivity.

There are a few important limitations to our study. In addition to the
observational nature of the cohort, the biology underlying the interaction
between TAGFA, insulin secretion, and insulin sensitivity is complex and
multifacetted {{?}} and as such there may be some residual confounding we were
unaware of or had not been able to adjust for. However, we have taken extensive,
empirically based precautions in identifying potential confounders and mediators
through the use of the DAG modeling, relying on previous literature, and through
QIC modeling fitting methods.

TAGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TAGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.
[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
control than 2hr-OGTT TAG

Since TAGFA were only quantified at the baseline visit and as such we cannot
investigate whether there are concomitant changes in TAGFA and the metabolic
measures over time. In relation to this limitation, while GEE is well suited to
longitudinal datasets, the complex biology and temporal nature of the
associations may restrict the ability of GEE to appropriately analyze and
interpret the data. However, to maximize the potential of GEE, we used the model
to infer that a given value of TAGFA could predict values of insulin sensitivity
or beta-cell function over a 6 year period. This in our view is a strength of
our analysis, as it reduces the chance of reverse causality given the tight
integration of the glucose and fatty acid metabolism pathways, as well as uses
GEE modeling appropriately {{? another word}}.

PLS is a well-established technique for constructing predictive models
of high dimensionality data structures (i.e. fatty acid composition), however a
limitation is that the initial models analyzed through PLS and the final
computed scores are not able to control for potential confounders and other
effect modifiers. Likewise, PLS is not able to handle longitudinal data so only
the baseline visit was used in the PLS analysis. In light of these limitations,
we still observe concordant results between the PLS and GEE analyses {{ add more }}.

Even with these limitations, our study also has several major strengths,
including the longitudinal design and the rigorous statistical techniques and
methods applied in analyzing the data. These statistical techniques are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, as well as having
both concentration and mol% data for the fatty acids.

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
advanced {{another word?}} statistical techniques designed to maximally utilize
the data. Two large prospective studies have been published that had similar
research objectives as the present analysis. One was a nested case-control
cohort (n=189 cases and n=189 controls) of the Framingham offspring study
[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and
less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
this study was the simpler statistical techniques used to analyze the data, 
which is not able to adequately use the longitudinal and high-dimensional 
dataset. {{ They're article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group, the authors report a 
positive correlation between total esterified (of which TAG make up the 
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a], similar as 
our results except for 18:1n-9.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a] as well as
studies showing higher refined carbohydrates (e.g. added sugars) and a higher
risk for diabetes {{cite}}. Our study extends these findings on DNL by showing
that TAGFA with 14 to 16 carbons all strongly predicted lower insulin
sensitivity. While these fatty acids also had a significant association with
beta-cell function, the results were not as strong as with the insulin
sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies have examined used markers
of DNL, such as the ratio between 18:2n-6 to 16:0 or 16:0 to 14:0 ratio
[@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can be created from DNL, they are also obtained from the diet [@Hodson2008a]. A
short, small feeding trial (n=24) was conducted to identify fatty acids that
most accurately reflected DNL, as potential biomarkers [@Lee2015a]. The study
found that palmitoleic acid (16:1n-7), directly measured DNL using isotopes, and
liver fat were all highly correlated with each other (r>0.50), suggesting that
16:1n-7 may be a good biomarker for hepatic DNL. In another small (n=14) feeding
trial, meal type (high fat vs low fat) was tested to determine its effect on DNL
and TAGFA composition [@Wilke2009a]. The authors reported that 14:0, 16:0, 16:1,
and 18:2 were higher in the low fat (high carbohydrate) group. These fatty acids
are similar to the fatty acids we found that clustered together, implicating 
these fatty acids as indicative of a higher carbohydrate diet. A higher
carbohydrate diet, particularly with simple carbohydrates, may lead to greater
DNL in an attempt to control blood glucose, increasing hepatic fat stores, and
increasing the amount of TAGFA in circulation. The increase in circulating 14
and 16 carbon fatty acids may then increase the exposure of tissues to lipotoxicity,
for instance from palmitic acid (16:0), which is well-known to have harmful
effects on tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedbacks involved. For instance, while greater DNL may promote muscle
resistance, the reverse may also be true [@Flannery2012a]. Higher insulin
resistance may encourage greater DNL to handle the higher blood glucose. To
illustrate, in a weight loss intervention trial (n=19 with TAGFA data),
participants who lost weight over 33 weeks showed higher insulin sensitivity and
lower TAGFA composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1,
etc) [@Schwab2008a]. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TAGFA more
strongly than TAGFA influencing insulin sensitivity will require complicated
research designs and analyses. Our own null results on this question may
tentatively point to an equal feedback mechansim between TAGFA and insulin
sensitivity.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(currently unpublished), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin
resistance. While NEFA may contribute substantially to the TAG fraction, there
may be something unique about the specific composition of the TAGFA fraction
that may allow it to enter insulin sensitive tissues more easily than NEFA. It
may be that VLDL receptors allow fatty acids to enter these tissue more easily than 
NEFA bound to albumin, or it may be due to the other influences on TAGFA
composition such as dietary intake.

@@

    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
    two groups: normogly and hypergly. results in general showed no diff between
    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
    not present results on whole fraction for IR)
{{ comment on unadjusted vs adjusted }}

{{ commment on null interactions with time }}

@@

Prospective cohorts:
    - Chien K, et al. A prediction model for type 2 diabetes risk among Chinese people. Diabetologia. 2009;52(3):443–450.
    - D’Agostino RB Jr, Hamman RF, Karter AJ, Mykkanen L, Wagenknecht LE, Haffner SM. Cardiovascular disease risk factors predict the development of type 2 diabetes: the insulin resistance atherosclerosis study. Diabetes Care. 2004;27(9):2234–2240.
    - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.
-->

>>>>

@@

We found that in an at-risk for diabetes Canadian population who were mainly female 
and of European ancestry, the specific TAGFA composition was strongly associated 
with insulin sensitivity and moderately associated with beta-cell function. In 
particular, TAGFA myristic acid (14:0), 7-tetradecenoic acid {{ confirm name }} 
(14:1n-7), palmitic acid (16:0), and palmitoleic acid (16:1n-7) all strongly and
negatively predicted lower insulin sensitivity. While most TAGFA were not
associated with beta-cell function, two fatty acids, 16:0 and 18:1n-7, were
associated negatively and positively, respectively, with both measures of 
beta-cell function. We also found that four TAGFA (14:0, 14:1n-7, 16:0, 16:1n-7)
clustered together, and that also strongly predicted lower insulin sensitivity. 
These four fatty acids are also involved in the DNL of refined and simple 
carbohydrates. Our results, which are in agreement with the current literature,
suggest that higher activity of DNL (potentially through higher intakes of
simple carbohydrates) may increase the risk for diabetes, likely through
worsening insulin sensitivity.

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using advanced
{{another word? appropriate?}} statistical techniques designed to maximally
utilize the data. Two large prospective studies have been published that had
similar research objectives as the present analysis. One was a nested
case-control cohort (n=189 cases and n=189 controls) of the Framingham offspring
study [@Rhee2015a], which found that TAGFA composition with a lower carbon chain
and less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
the study was the simpler statistical techniques used to analyze the data, 
which are not able to adequately use the longitudinal and high-dimensional 
dataset. {{ They're article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group [@Kotronen2009a], the authors
report a positive correlation between total esterified (of which TAG make up the
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR, similar as our results except
for the 18:1n-9 finding.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]. Our study
extends these findings on DNL by showing that TAGFA with 14 to 16 carbons all
strongly predicted lower insulin sensitivity. While these fatty acids also had a
significant association with beta-cell function, the results were not as strong
as with the insulin sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A short, small feeding trial (n=24) was conducted to identify
fatty acids that most accurately reflected DNL, as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TAGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet,
particularly with simple carbohydrates, may lead to greater DNL in an attempt to
control blood glucose, increasing hepatic fat stores, and increasing the amount
of TAGFA in circulation. The higher concentration of circulating 14 and 16
carbon fatty acids may then expose tissues to greater lipotoxicity, for instance
from palmitic acid (16:0), which is well-known to have harmful effects on
tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedback mechanisms involved. For instance, while greater DNL may
promote muscle insulin resistance, the reverse may also be true
[@Flannery2012a]. Higher insulin resistance may encourage greater DNL to handle
the higher blood glucose. To illustrate this, in a weight loss intervention trial
(n=19 with TAGFA data), participants who lost weight over 33 weeks showed higher
insulin sensitivity and lower TAGFA composition indicative of lower DNL (less
14:0, 14:1, 16:0, 16:1, etc) [@Schwab2008a]. Given the complex biological
mechanisms and feedback loops involved, disentangling whether insulin
sensitivity influences TAGFA more strongly than TAGFA influencing insulin
sensitivity will require more complicated research designs and analyses. Our own
null results on this question may tentatively point to an equally influential
feedback mechanism between TAGFA and insulin sensitivity.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during the fed state while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(currently unpublished), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin 
resistance. While NEFA may contribute substantially to the TAG fraction, there 
may be something unique about the specific composition of the TAGFA fraction 
that may allow it to enter insulin sensitive tissues more easily than NEFA. It 
may be that VLDL receptors allow fatty acids to enter these tissue more easily
than NEFA bound to albumin, or it may be due to the other influences on TAGFA 
composition such as dietary intake.

@@

- Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
two groups: normogly and hypergly. results in general showed no diff between
groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
not present results on whole fraction for IR)

{{ comment on unadjusted vs adjusted }}

- Potential clinical application

@@

There are a few important limitations to our study. In addition to the
observational nature of the cohort, the biology underlying the interaction
between TAGFA, insulin secretion, and insulin sensitivity is complex and
multifaceted and as such there may be some residual confounding we were
unaware of or had not been able to adjust for. However, we have taken extensive,
empirically based precautions in identifying potential confounders and mediators
through the use of the DAG modeling, relying on previous literature, and through
QIC model fit comparison methods.

<!--
TAGFA can fluctuate substantially throughout the day and to control for this all
PROMISE participants came for their clinic visits in the morning in a fasted
state. Some of the inherent variability in TAGFA fluctuation can be managed by
standardizing the data collection time, though there may still be some
{{residual error? random error?}}.
[@Rhee2011a] fasting TAG status better as discriminating between DM case vs
control than 2hr-OGTT TAG
-->

TAGFA were only quantified at the baseline visit and as such we cannot
investigate whether there are concomitant changes in TAGFA and the metabolic
measures over time. In relation to this limitation, while GEE is well suited to
longitudinal datasets, the complex biology and temporal nature of the
associations may restrict the ability of GEE to appropriately analyze and
interpret the data. However, to maximize the potential of GEE, we used the model
to infer that a given value of TAGFA could predict values of insulin sensitivity
or beta-cell function over a 6 year period. This in our view is a strength of
our analysis, as it reduces the chance of reverse causality given the tight
integration of the glucose and fatty acid metabolism pathways, as well as
maximizes the specific usage of the GEE modeling.

PLS is a well-established technique for constructing predictive models of high
dimensionality data structures (i.e. fatty acid composition), however a 
limitation is that the initial models analyzed through PLS and the final 
computed scores are not able to control for potential confounders and other 
effect modifiers. Likewise, PLS is not able to handle longitudinal data so only 
the baseline visit was used in the PLS analysis. In light of these limitations, 
we still observe concordant results between the PLS and GEE analyses.

Even with these limitations, our study has several notable strengths, including
the longitudinal design and the rigorous statistical techniques and methods
applied in analyzing the data. These statistical techniques take advantage of
the longitudinal data to allow appropriate investigation of temporal
relationships and are able to handle the multivariate nature of the data.
Lastly, our cohort contains highly detailed and comprehensive variable
measurements at each collection visit, as well as having both concentration and
mol% data for the fatty acids.
